---
title: "ballondor2"
author: "Ahmad Khan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load libraries
library(tidyverse)
library(broom)
library(ggplot2)
library(caret)
library(glmnet)
library(rpart)
library(rpart.plot)

# Read and fix column names
bdr <- read_csv("Ballon D'or Master 3.csv")

# Replace spaces and dots in column names with underscores
names(bdr) <- names(bdr) %>%
  str_replace_all(" ", "_") %>%
  str_replace_all("\\.", "_")

# Check cleaned names
glimpse(bdr)

# Make sure Position is a factor
bdr <- bdr %>% mutate(Position = as.factor(Position)) 


```



```{r}
# Split by position
bdr_goalkeeper <- filter(bdr, Position == "Goalkeeper")
bdr_defender   <- filter(bdr, Position == "Defender")
bdr_midfielder <- filter(bdr, Position == "Midfielder")
bdr_forward    <- filter(bdr, Position == "Forward")

# Goalkeeper model
model_goalkeeper <- lm(Pct_of_Possible ~ Appearances + Major_Cups + Minor_Cups + Clean_Sheets + Major_Awards, data = bdr_goalkeeper)

# Defender model
model_defender <- lm(Pct_of_Possible ~ Major_Cups + Clean_Sheets + Major_Awards, data = bdr_defender)

# Midfielder model
model_midfielder <- lm(Pct_of_Possible ~ Assists + Major_Cups + Major_Awards, data = bdr_midfielder)

# Forward model
model_forward <- lm(Pct_of_Possible ~ Goals + Assists + Major_Cups + Major_Awards, data = bdr_forward)

# Summarize
summary(model_goalkeeper)
summary(model_defender)
summary(model_midfielder)
summary(model_forward)



```
```{r}
plot(model_goalkeeper)
```

```

```{r}
# Prepare data: Select numeric features
bdr_model_data <- bdr %>%
  dplyr::select(Pct_of_Possible, Goals, Assists, Appearances, Clean_Sheets, Major_Cups, Minor_Cups, Major_Awards)

# Remove NA rows if any
bdr_model_data <- na.omit(bdr_model_data)

# Create X and y
X <- bdr_model_data %>% dplyr::select(-Pct_of_Possible) %>% as.matrix()
y <- bdr_model_data$Pct_of_Possible


# Split into train and test sets
set.seed(123)
train_idx <- sample(1:nrow(X), 0.8 * nrow(X))
X_train <- X[train_idx, ]
y_train <- y[train_idx]
X_test  <- X[-train_idx, ]
y_test  <- y[-train_idx]

# Fit Elastic Net
cv_model <- cv.glmnet(X_train, y_train, alpha = 0.5)  # alpha = 0.5 => Elastic Net

# Best lambda
best_lambda <- cv_model$lambda.min

# Final model
elastic_model <- glmnet(X_train, y_train, alpha = 0.5, lambda = best_lambda)

# Predict on test set
y_pred <- predict(elastic_model, X_test)

# Evaluate RMSE
rmse <- sqrt(mean((y_test - y_pred)^2))
print(paste("Elastic Net Test RMSE:", round(rmse, 2)))


```

```{r}
# Unified model with Position as factor
ancova_model <- lm(Pct_of_Possible ~ Goals + Assists + Appearances + Clean_Sheets + Major_Cups + Minor_Cups + Major_Awards + Position, data = bdr)

summary(ancova_model)

```



```{r}
cor(bdr[, c("Pct_of_Possible", "Goals", "Assists", "Appearances", "Clean_Sheets", "Major_Cups", "Minor_Cups", "Major_Awards")])
```
```{r}
ancova_interaction <- lm(Pct_of_Possible ~ (Goals + Assists + Appearances + Clean_Sheets + Major_Cups + Minor_Cups + Major_Awards) * Position, data = bdr)
anova(ancova_model, ancova_interaction)
```

```{r}
library(car)
vif(ancova_model)
```

```{r}
plot(ancova_model)
```
```{r}
# Load required libraries
library(ggplot2)
library(broom)

# Get residuals and fitted values from the model
model_data <- augment(ancova_model)

# Residuals vs Fitted Plot
ggplot(model_data, aes(.fitted, .resid)) +
  geom_point() +
  geom_smooth(method = "loess", col = "red", se = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  labs(title = "Residuals vs Fitted", x = "Fitted Values", y = "Residuals") +
  theme_bw()

# Q-Q Plot
ggplot(model_data, aes(sample = .std.resid)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "Q-Q Plot of Residuals", x = "Theoretical Quantiles", y = "Standardized Residuals") +
  theme_bw()
```

```{r}
# Load required libraries
library(car)
library(flextable)

# Calculate GVIF
gvif_raw <- vif(ancova_model)

# Convert `gvif_raw` to a properly structured data frame
gvif_data <- data.frame(
  Variable = rownames(gvif_raw),                     # Variable names
  Scaled_GVIF = gvif_raw[, "GVIF^(1/(2*Df))"],       # Scaled GVIF values
  row.names = NULL                                   # Remove row names
)

# Create a flextable for the GVIF table
gvif_table <- flextable(gvif_data) %>%
  set_caption("Generalized Variance Inflation Factor (GVIF)") %>%
  colformat_double(j = "Scaled_GVIF", digits = 4) %>%
  italic(j = 1, part = "header") %>%
  autofit()

# Print the table
print(gvif_table)
```

```{r}
# Load required libraries
library(car)
library(ggplot2)

# Calculate GVIF
gvif_raw <- vif(ancova_model)

# Convert `gvif_raw` to a data frame
gvif_data <- data.frame(
  Variable = rownames(gvif_raw),                     # Variable names
  Scaled_GVIF = gvif_raw[, "GVIF^(1/(2*Df))"]        # Scaled GVIF values
)

# Plot the Scaled GVIF values
ggplot(gvif_data, aes(x = reorder(Variable, Scaled_GVIF), y = Scaled_GVIF, fill = Scaled_GVIF)) +
  geom_bar(stat = "identity", color = "black") +
  geom_hline(yintercept = 2, linetype = "dashed", color = "red", size = 0.8) +
  labs(
    title = "Scaled GVIF Values for Model Variables",
    x = "Variable",
    y = "Scaled GVIF",
  ) +
  scale_fill_gradient(low = "lightblue", high = "steelblue") +
  coord_flip() +  # Flip coordinates for better readability
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 13),
    legend.position = "none"
  )
```

```{r}
# Perform ANOVA comparison
anova_table <- anova(ancova_model, ancova_interaction)

# Convert ANOVA output to a data frame
anova_df <- as.data.frame(anova_table) %>%
  mutate(
    `Model` = c("Full Model", "Model+Interaction"),
    `P-value` = ifelse(`Pr(>F)` < 2e-16, "< 2e-16", format(`Pr(>F)`, scientific = TRUE, digits = 2))
  ) %>%
  select(Model, Res.Df, RSS, Df, `Sum of Sq`, F, `P-value`)  # Reorder columns

# Create the flextable
anova_table_flex <- flextable(anova_df) %>%
  set_caption("ANOVA Table: Testing Significance of Interaction Terms") %>%
  colformat_double(j = c("RSS", "Sum of Sq", "F"), digits = 4) %>%
  italic(j = 1, part = "header") %>%
  autofit()

# Print the table
print(anova_table_flex)
```

```{r}
# Train Decision Tree
tree_model <- rpart(Pct_of_Possible ~ Goals + Assists + Appearances + Clean_Sheets + Major_Cups + Minor_Cups + Major_Awards + Position,
                    data = bdr, method = "anova")

# Plot Tree
rpart.plot(tree_model, type = 2, extra = 101, under = TRUE, faclen = 0, cex = 0.8)

```
```{r}
# Diagnostics for ANCOVA model
par(mfrow = c(2,2))
plot(ancova_model)

# Variable Importance (Elastic Net)
coef_elastic <- as.vector(coef(elastic_model))
names(coef_elastic) <- c("(Intercept)", colnames(X_train))
coef_elastic_df <- data.frame(Feature = names(coef_elastic), Coefficient = coef_elastic)

coef_elastic_df %>%
  filter(Feature != "(Intercept)") %>%
  ggplot(aes(x = reorder(Feature, Coefficient), y = Coefficient)) +
  geom_col() +
  coord_flip() +
  labs(title = "Elastic Net Feature Importance", x = "", y = "Coefficient")

```

```{r}
# Start
bdr_preds <- bdr

# --- Predict Position models separately ---
bdr_preds <- bdr_preds %>%
  mutate(
    Pred_GK  = if_else(Position == "Goalkeeper", predict(model_goalkeeper, newdata = bdr_preds), NA_real_),
    Pred_DF  = if_else(Position == "Defender", predict(model_defender, newdata = bdr_preds), NA_real_),
    Pred_MF  = if_else(Position == "Midfielder", predict(model_midfielder, newdata = bdr_preds), NA_real_),
    Pred_FW  = if_else(Position == "Forward", predict(model_forward, newdata = bdr_preds), NA_real_)
  ) %>%
  mutate(
    Pred_Position_Model = coalesce(Pred_GK, Pred_DF, Pred_MF, Pred_FW)
  )

# --- Predict ANCOVA model ---
bdr_preds$Pred_ANCOVA <- predict(ancova_model, newdata = bdr_preds)

# --- Predict Elastic Net model ---
X_full <- bdr %>%
  dplyr::select(Goals, Assists, Appearances, Clean_Sheets, Major_Cups, Minor_Cups, Major_Awards) %>%
  as.matrix()

bdr_preds$Pred_ENET <- as.numeric(predict(elastic_model, newx = X_full, s = best_lambda))  # <-- force to numeric

```

```{r}
# Select only important columns and save
bdr_preds %>%
  dplyr::select(
    Year, Player, Position, Place, Points, Pct_of_Possible,
    Pred_Position_Model, Pred_ANCOVA, Pred_ENET
  ) %>%
  write_csv("bdr_predictions.csv")


```

```{r}
library(flextable)
library(broom)
library(dplyr)

print_model_summary <- function(model, model_name = "Model", newdata = NULL, interval_type = c("confidence", "prediction")) {
  # 1. Coefficient Table
  model_summary <- tidy(model)
  
  coef_table <- flextable(model_summary) %>%
    set_caption(paste("Coefficient Estimates:", model_name)) %>%
    colformat_double(j = c("estimate", "std.error", "statistic"), digits = 4) %>%
    italic(j = 1, part = "header") %>%
    compose(
      j = "p.value",
      value = as_paragraph(
        ifelse(model_summary$p.value < 2e-16,
               "< 2e-16",
               format(model_summary$p.value, scientific = TRUE, digits = 2))
      )
    ) %>%
    autofit()
  
  print(coef_table)
  
  # 2. Model Performance Statistics Table
  model_glance <- glance(model)
  
  perf_df <- data.frame(
    `R-squared` = round(model_glance$r.squared, 4),
    `Adj. R-squared` = round(model_glance$adj.r.squared, 4),
    `Residual Std. Error` = round(model_glance$sigma, 4),
    `F-statistic` = round(model_glance$statistic, 4),
    `F-stat p-value` = ifelse(model_glance$p.value < 2e-16, "< 2e-16",
                              format(model_glance$p.value, scientific = TRUE, digits = 2)),
    `DF` = paste0(model_glance$df[1], " & ", model_glance$df[2])
  )
  
  perf_table <- flextable(perf_df) %>%
    set_caption(paste("Model Fit Statistics:", model_name)) %>%
    autofit()
  
  print(perf_table)

  # 3. Optional: Confidence or Prediction Intervals
  if (!is.null(newdata)) {
    for (type in interval_type) {
      pred <- predict(model, newdata = newdata, interval = type)
      pred_df <- as.data.frame(pred)
      
      pred_table <- flextable(pred_df) %>%
        set_caption(paste0(toupper(substring(type, 1, 1)), substring(type, 2), 
                           " Interval for ", model_name)) %>%
        colformat_double(digits = 4) %>%
        autofit()
      
      print(pred_table)
    }
  }
}


```


```{r}
# For full model
print_model_summary(ancova_model, model_name = "ANCOVA Model")

# For goalkeeper model
print_model_summary(model_goalkeeper, model_name = "Goalkeeper Full Model")

print_model_summary(model_midfielder, model_name = "Midfielder Reduced Model")

# If you want to add predictions for a new county (or new observation)
new_data <- data.frame(
  mean_age = 45,
  prop_male = 0.48,
  prop_educ_4yr = 0.25,
  prop_white = 0.70,
  mean_ftotinc = 50000
)
print_model_summary(ancova_model, model_name = "Full Model", newdata = new_data, interval_type = c("confidence", "prediction"))

```


